Visual diagnostics for constrained optimisation with application to guided tours

### 2020-10-13 paper/paper.Rmd
tuning languages here and there and everywhere

- name: Author 1
  thanks: The authors gratefully acknowledge ...
  affiliation: Department of YYY, University of XXX
- name: Author 2
  affiliation: Department of ZZZ, University of WWW
keywords:
- optimisation
- projection pursuit 
- guided tourr
- visual 
- diagnostics 
- R
abstract: Friedman & Tukey commented on their initial paper on projection pursuit in 1974 that "the  technique  used  for  maximising  the  projection index strongly influences both the statistical and the computational aspects of the procedure." While many projection pursuit indices have been proposed in the literature, few concerns the optimisation procedure. In this paper, we developed a system of diagnostics aiming to visually learn how the optimisation procedures find its way towards the optimum. This diagnostic system can be applied more generally to help practitioner to unveil the black-box in randomised iterative (optimisation) algorithms. An R package, ferrn, has been created to implement this diagnostic system.
bibliography: biblio.bib
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont, bbm, array, booktabs}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 3
editor_options: 
  chunk_output_type: console
---

@@

```{r external, include = FALSE, cache = FALSE}
library(knitr)
read_chunk(here::here('scripts/vis-diag.R'))
read_chunk(here::here('scripts/examples.R'))
read_chunk(here::here('scripts/kol-better-default.R'))
read_chunk(here::here('scripts/kol-better-random-default.R'))
read_chunk(here::here('scripts/kol-better-random-tuned.R'))
read_chunk(here::here('scripts/implementation.R'))
```

@@

<!-- We can see from Figure \ref{tour-path} that only the data projected using the accepted and interpolated bases are animated by tour; none of the searching bases sampled by the optimsiation algorithms are presented in the tour animation. Although these algorithms have been demonstrated to find the global maximum in various literature as we cited above, they focus solely on finding the global maximum and fewer tools are available to explore how each algorithm searches the parameter space and compare the results between different algorithms.  -->

<!-- This study is meaningful because **as more complex index functions are proposed based on statistical theory, only computationally effective optimisation algorithms allow us to find the projection basis that presents interesting structure in the data**. -->

<!-- Hence the study **presents a gallery of diagnostic plots applicable to optimisation algorithms.** -->

Visualisation has been widely used in exploratory data analysis. Presenting information in a graphical format often allows people to see information they would otherwise not see. This motivates our work of creating plots to diagnose optimisation algorithms in the context of projection pursuit guided tour, with the aim to understand and compare features of different existing algorithms.

In an optimization problem the goal is to find the best solution within the space of all feasible solutions which typically is represented by a set of constraints. The problem consists in optimizing an objective function $f: S \rightarrow \mathbb{R}$ with $S \in \mathbb{R}^n$ in a reduced space given by the problem constraints to either minimize or maximize a function.

Projection pursuit and guided tour are exploratory data analysis tools that detect interesting structure of high dimensional data through projection on low dimensional space. Optimisation is applied here to search for the low dimensional space that finds the most interesting projection.

The remainder of the paper is organised as follows. 
Section \ref{optim} provides a literature review of optimisation methods, specifically the line search methods used in projection pursuit guided tour. 
Section \ref{tour} reviews projection pursuit guided tour, forms the optimisation problem, and introduces three main existing algorithms.
Section \ref{vis-diag} presents the new visual diagnostics design, from forming the data object to the definition of different diagnostic plots with some small examples. 
Section \ref{application} shows the application of how the diagnostic plots designed in section \ref{vis-diag} can be used to understand and compare different algorithms and how they contribute to modifications that improve the algorithms.
Finally, Section \ref{implementation} describes the R package: ferrn, that implements all the visual diagnostics above.

@@

Given an optimisation problem, two basic approaches find the optimum based on different thinking.
An analytical approach aims to find the optimal solution in a finite number of steps, but a potential issue with it is that the closed-form solution may not be available when the problem starts to become complex. 
An iterative approach, on the other hand, finds the optimum based on the idea of making progressive improvement to the current solution. An iterative method may end up finding a local optimum but the progressive nature of the algorithm allows the practitioner to decide when to stop if a desirable accuracy has been achieved.

A traditional while often used in practice is an iterative method called *line search method* [@fletcher2013practical]. In a simple one-dimensional problem of finding the $x$ that minimises $f(x)$, line search achieves the goal via an iterative algorithm in the form of Equation \ref{eq:line-search}.

@@

where $d^{(j)}$ is the searching direction in iteration $j$, and $\alpha_j$ is the step-size. Strictly speaking, $\alpha_k$ is chosen by another minimisation of $f(x^{(j)} + \alpha* d^{(j)})$ with respect to $\alpha$ and theoretical results have demonstrated the global convergence of the algorithm when the exact minimisation of $\alpha_j$ is attained [@curry1944method]. In practice, this second minimisation is rarely implemented due to its computational demanding or even the existence of such a minimisation. A more realistic approach is to impose a mandatory decrease in the objective function for each iteration: $f^{(j+1)}> f^{(j)}$ and despite we lose the guarantee on global convergence, this approach turns out to be efficient in practical problems.

@@

Modern development of the line search methods focuses on proposing different computation on the searching direction: $d^{(j)}$ and various approximations on the step size: $\alpha_j$ catered for practical optimisation problems. The specific problem context we are interested in is called projection pursuit guided tour. 
Projection pursuit and guided tour are two separate methods in exporatory data analysis focusing on different aspects: coined by @friedman1974projection, projection pursuit detects interesting structures (i.e. clustering, outliers and skewness) in multivariate data via low dimensions projection; whilst guided tour is a particular variation in a broader class of data visualisation method called tour.

Let $\mathbf{X}_{n \times p}$ be the data matrix, an n-d projection can be seen as a linear transformation $T: \mathbb{R}^p \mapsto \mathbb{R}^d$ defined by $\mathbf{P} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{P}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis. Define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projection basis $\mathbf{A}$ onto an index value $I$, this function is commonly known as the projection pursuit index (PPI) function, or the index function and is used to measure the "interestingness" of a projection. A number of index functions have been proposed in the literature to detect different data structures, including Legendre index [@friedman1974projection], Hermite index [@hall1989polynomial], natural Hermite index [@cook1993projection], chi-square index [@posse1995projection], LDA index [@lee2005projection] and PDA index [@lee2010projection].

In their initial paper, @friedman1974projection noted that "..., the technique used for maximising the projection index strongly influences both the statistical and the computational aspects of the procedure." Hence, effective optimisation algorithms are necessary for projection pursuit to find the bases that give interesting projections. While we leave the formal construction of the optimisation problem and existing algorithms to section \ref{tour-optim}, we outline the general idea here. Given a random starting (current) basis, projection pursuit repeatedly searches for candidate bases nearby until it finds one with higher index value than the current basis. In the second round, that basis becomes the current basis and the repetitive sampling  continues. The process ends until no better basis can be found or one of the termination criteria is reached.

Before introducing the guided tour, we shall be familiar with the general tour method [@cook2008grand]. A tour produces animated visualisation of the high dimensional data via rotating low dimension planes. The smoothness of the animation is ensured by computing a series of intermediate planes between two low dimension planes via geodesic interpolation and we refer readers to @buja2005computational for the mathematical details. Iteratively choosing different low dimension planes and interpolating between them forms a tour path. Different types of tour methods choose the low dimensional planes differently and we mention two other types of tour that are commonly used. A grand tour selects the planes randomly in the high dimensional space and hence serves as an initial exploration of the data. Manual control allows researches to fine-tuning an existing projection by gradually phase in and out one variable.

Guided tour chooses the planes produced by optimising the projection pursuit index function. Figure \ref{tour-path} shows a sketch of the tour path consisting of the blue frames produced by the projection pursuit optimisation algorithm iteratively and the white frames, which are the interpolations between two blue frames. The tour method has been implemented in the *tourr* package in R, available on the Comprehensive R Archive Network at [https://cran.r-project.org/web/packages/tourr/](https://cran.r-project.org/web/packages/tourr/) [@wickham2011tourrpackage].

@@

(ref:tour-path) Each square (frame) represents the projected data with a corresponding basis. Blue frames have bases found by an optimisation algorithm iteratively whilst the white frames are constructed by geodesic interpolation between two blue frames.

<!-- The pseudo-code below illustrates the implementation of guided tour in the tourr package. Given an projection pursuit index function and a randomly generated projection basis (current basis), the optimisation procedure produces a target basis inside `generator()`  Both the current basis and the target basis will be supplied to `tour_path()` to prepare information needed for constructing a geodesic path. This information is then used to compute a series of interpolating bases inside the `tour()` function. All the basis will be sent to create animation for visualising the tour in the `animate()` function.   -->

<!-- ```{r algo,eval = FALSE, echo = TRUE} -->
<!-- animation <- function(){ -->

<!--   # compute projection basis  -->
<!--   tour <- function(){ -->

<!--     # construct bases on the tour path -->
<!--     new_geodesic_path <- function(){ -->
<!--       tour_path <- function(){ -->

<!--         # GENERATOR: generate projection basis via projection pursuit -->
<!--         guided_tour <- function(){ -->
<!--           generator <- function(){ -->

<!--             # define projection pursuit index -->
<!--             # generate the target basis from the current basis via optimisation -->
<!--           } -->
<!--         } -->

<!--         # prepare geodesic information needed for interpolating along the tour path -->
<!--       } -->
<!--     } -->

<!--     # INTERPOLATOR: interpolate between the current and target basis  -->
<!--     function(){ -->
<!--       # generate interpolating bases on the geodesic path -->
<!--     } -->
<!--   } -->

<!--   # animate according to different display methods -->
<!-- } -->
<!-- ``` -->

@@

Now we begin to formulate the optimisation problem. Given a randomly generated starting basis $\mathbf{A}_1$, projection pursuit finds the final projection basis $\mathbf{A}_T$ that satisfies the following optimisation problem:

@@

where $I_d$ is the d-dimensional identity matrix. The constraint requires the projection basis $\mathbf{A}$ to be an orthogonal matrix with each column vector being orthonormal.

There are several features of this optimisation that are worth noticing. First of all, it is a multivariate constraint optimisation problem. Since the decision variables are the entries of a projection basis, it is required to be orthonormal. It is also likely that the objective function is non-differentiable or the gradient information is simply not available. In this case, we will need to either use some approximation of the gradient or turn to derivative free methods. Given the goal of projection pursuit as finding the basis with the largest index value, the optimisation problem needs to be able to find the global maximum. Along the way, local maximum may also be of our interest since they could present unexpected interesting projections. There is also one computational consideration: the optimisation procedure needs to be easy to compute since the tour animation is played in real-time.

<!-- - *constraint optimisation*: our problem requires the project basis to be orthonormal -->

<!-- - *Multivariate optimisation*: In projection pursuit, the decision variable includes all the entries in the projection matrix, which is high-dimensional. Researchers would be better off if they could understand the relative position of different projection matrix in the high-dimensional space. -->

<!-- - *optimising non-smooth function*: When the objective function is non-differentiable, derivative information can not be obtained, which means traditional gradient- or Hessian- based methods are not feasible. Stochastic optimisation method could be an alternative to solve these problems. (derivative information is not available) -->

<!-- - *Finding global maximum*: Although finding local maximum is relatively easy with developed algorithms, it is generally hard to guarantee global maximum in a problem where the objective function is complex or the number of decision variables is large. Also, there are discussions on how to avoid getting trapped in a local optimum in the literature. (the exact $\alpha$ that minimised $f(x^{(k)}+ \alpha d^{(k)})$ can be hard to find or sometiems even impossible, imposing certain decrease, how every this doesn't guarantee global convergence) -->

<!-- - *computation speed*: The optimisation procedure needs to be fast to compute since the tour produces real-time animation of the projected data. -->

<!-- - *consistency result in stochastic optimisation*: In stochastic algorithm, researchers usually set a seed to ensure the algorithm produces the same result for every run. This practice supports reproducibility, while less efforts has been made to guarantee different seeds will provide the same result.  -->

<!-- - *role of interpolation in PP optimisation*: An optimisation procedure usually involves iteratively finding projection bases that maximise the index function, while tour requires geodesic interpolation between these bases to produce a continuous view for the users. It would be interesting to see if the interpolated bases could, in reverse, help the optimisation reach faster convergence. -->

<!-- the interpolation can be seen as giving information on the direction of the derivative. -->

<!-- *Think about how does your package help people to understand optimisation* -->

<!--  - diagnostic on stochastic optim -->
<!--  - vis the progression of multi-parameter decision variable  -->
<!--  - understanding learning rate - neighbourhood parameter -->
<!--  - understand where the local & global maximum is found - trace plot - see if noisy function -->

@@

Below we introduce three possible algorithms: `search_better` and `search_better_random` are derivative free methods that sample candidate bases in the neighbourhood, whilst `search_geodesic` is an analogue of gradient ascent on the manifold of the projection basis.

\begin{algorithm}
\SetAlgoLined
  \SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{$\mathbf{A}_{\text{cur}}$, $f$, $\alpha$, $l_{\max}$} 
    \Output{$\mathbf{A}_{l}$}
  initialisation\;
  Set $l = 1$\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ and orthogonise $\mathbf{A}_{l}$\;
    Compute $I_{l}  = f(\mathbf{A}_{l})$\;
    \If{$I_{l} > I_{\text{cur}}$}{
      \KwRet{$\mathbf{A}_{l}$} \;
      }
    $l = l + 1$\;
  }
  \caption{random search}
  \label{random-search}
\end{algorithm}

`search_better` is a random search device that samples a candidate basis $\mathbf{A}_{l}$ in the neighbourhood of the current basis $\mathbf{A}_{\text{cur}}$ by $\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ where $\alpha$ controls the radius of the sampling neighbourhood and $\mathbf{A}_{\text{rand}}$ is a randomly generated matrix with the same dimension as $\mathbf{A}_{\text{cur}}$. $\mathbf{A}_{l}$ is then orthogonalised to ensure the orthonormal constraint is fulfilled. When a basis is found with index value higher than the current basis $\mathbf{A}_{\text{cur}}$, the search terminates and outputs the basis for guided tour to construct an interpolation path. The next iteration of search begins after adjusting $\alpha$ by a cooling parameter: $\alpha_{j+1} = \alpha_j * \text{cooling}$. Another termination condition is when the maximum number of iteration $l_{\max}$ is reached. A slightly different cooling scheme has been proposed by @posse1995projection to include a halving parameter $c$. Rather than reducing the radius of the searching neighbourhood, $\alpha$, at each iteration, Posse's design only adjust $\alpha$ if the last search takes more than $c$ times to find an accepted basis to avoid the searching space being reduced too fast. The algorithm of `search_better` is summarised in Algorithm \ref{random-search}.  [mention orthonormalise to ensure the constraint is fulfilled; don't use derivative information but a random search]

@@

Simulated annealing [@kirkpatrick1983optimization, @bertsimas1993simulated] uses the same sampling process as `search_better` but allow a probabilistic acceptance of a basis with lower index value based on a cooling scheme $T(l)$. Given an initial $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{\log(l + 1)}$. When a candidate basis fails to have an index value larger than the current basis, simulated annealing gives it a second chance to be accepted with probability $$P= \min\left\{\exp\left[-\frac{I_{\text{cur}} - I_{l}}{T(l)}\right],1\right\}$$ where $I_{(\cdot)}$ denotes the index value of a given basis. This implementation allows the algorithm to jump out of a local maximum and enables a more holistic search of the whole parameter space. This feature is particularly useful when the dimension of the projected space is smaller than the number of informative variables in the dataset (i.e. a one dimensional projection of the dataset with two informative variables). The algorithm can be written as replacing line 5-8 of Algorithm \ref{random-search} with Algorithm \ref{simulated_annealing}.

@@

@cook1995grand used a gradient ascent algorithm on the manifold of the projection bases. In gradient ascent, one first find the direction for improvment via computing the gradient information. In `search_geodesic`, $2n$ bases are first generated in a tiny neighbourhood of the current basis, controlled by the neighbourhood parameter $\delta$. A geodesic is then constructed using the current basis and one of the $2n$ bases that have the highest index value. If the neighbourhood parameter $\delta$ is tiny, the geodesic constructed is an analogue of the gradient information in the curved space and can work as the searching direction. In gradient ascent, the next step is to conduct a line search to find the best improvement along the gradient direction. In `search_geodesic`, this step is replaced by optimising the index value along the geodesic direction constructed before over an 90 degree angle from $-\pi/4$ to $\pi/4$. The optima $\mathbf{A}_{**}$ is outputted for the current iteration if the percentage change in the index value between $\mathbf{A}_{**}$ and $\mathbf{A}_{\text{cur}}$ is greater than a threshold value. As above, another termination condition is when $l_{\max}$ is reached.  Algorithm \ref{search-geodesic} summarises the steps in geodesic search.

@@

To be able to make diagnostic plot, the optimisation algorithm should populate a data structure that contains the key elements of the algorithm. When the algorithm runs, key information regarding the decision variable, objective function and hyper-parameters needs to be recorded and stored as a data object for future analysis.

@@

<!-- [Do I need this paragraph to details the grammar of graphic? xxx - don't think so; not particularly relevant - assume most people know it roughly] -->
<!-- A layer includes 1) the data that powers the plot; 2) a geometric object that represents the visual shape of the data and 3) relevant statistical transformation that transform the information in the data to the information used to draw the geometric object. An important concept in the grammar of graphic is *aesthetic mapping*. Aesthetic mapping links the variable in a dataset to information needed to produce a geometric object. For example, we map one variable on the x-axis and another on the y-axis to create a scatterplot. Another example of creating a boxplot involves 1) mapping one variable on the x-axis and 2) transforming another variable to its five-point-summary and mapping the five-point-summary to the y-axis.  -->

In the optimisation algorithms for projection pursuit, three main elements to record are 1) projection bases: $\mathbf{A}$, 2) index values: $I$, and 3) State: $S$, which labels the observation with detailed stage in the optimisation. Multiple iterators are also  needed to index the data collected at different levels. $t$ is a unique identifier that prescribes the natural ordering of each observation; $j$ is the counter for each search-and-interpolate round, which remains the same within one round and has an increment of one once a new round starts. $l$ is the counter for each search/interpolation allowing us to know how many basis the algorithm has searched before finding one to output. There are other parameters that are of our interest and we denote them as *$V_{p}$*. In projection pursuit, this includes $V_1 = \text{method}$, which tags the name of the algorithm used and $V_2 = \text{alpha}$, the neighbourhood parameter that controls the size in sampling candidate bases.  A matrix notation of the data structure is presented in Equation \ref{eq:data-structure}.

@@

where $T^{\prime} = T + k_{J}+ l_{J+1}$. Note that we deliberately denote the last round of search as $j = J+1$ and in that round there is no output/interpolation basis and the algorithm terminates. This notation allows us to denote the last complete search-and-interpolate round as round $J$ and hence the final basis is $A_T$ and highest index value found is $I_T$.

@@

It is worth noticing that the data structure constructed above meets the tidy data principle [@wickham2014tidy] that states

1) each observation forms a row, 
2) each variable forms a column, and 
3) each type of observational unit forms a table

The wrangling and visualisation of tidy data have been greatly simplified by the well-known dplyr[@dplyr] and ggplot2[@ggplot2] package.

With a constructed data object, the construction of diagnostic plots is inspired by the concept of grammar of graphic [@wickham2010layered], which powers the primary graphical system in R, ggplot2 [@ggplot2]. In grammar of graphic, plots are not defined by its appearance (i.e. boxplot, histogram, scatter plot etc) but by "stacked layers". Using this design, ggplot does not have to develop a gazillion of functions that each produces a different type of plot from a different data structure. Instead, it aesthetically maps variables (and its statistical transformation) in a dataset to different geometric objects (points, lines, box-and-whisker etc) and builds the plot through overlaying different layers.

<!-- The plot design of visual diagnostics depends on the characteristics of the variables to plot. If the searching points are of interest [point geom is not good; need summarise. While updating points are more manageable -> point geom]. However, one must realise that with hundreds or thousands of searching points, exploring the sample space in animation could be slow and this is because of the time it takes to render hundreds point. [stratefy may help? - may need more work here.] -->

<!-- ### Explore the value of objective function {#static} -->

<!-- In tour, all the points recorded in the global object can be roughly divided into two broad categories: searching points and interpolating points -->

<!-- - *Searching points* include the observations recorded in the searching algorithm to find the target basis. The points for target bases is also included in the searching points and there is one such point per `tries`. -->

<!-- - *interpolating points* exist in the guided tour to produce continuous animated view from one target basis to another and it doesn't have `loop` value. -->

<!-- In the current implementation of the `tourr` package, while the target basis generated by the projection pursuit can be accessed later via `save_history()`, interpolating bases and those randomly nearby bases generated in the optimisation are not stored. This creates difficulties for fully understand the behaviour of the optimisation and interpolation of tour in complex scenario **[need a rephrase this part].**  -->

@@

```{r toy-pca-aniamted, fig.cap = "(ref:toy-pca-aniamted)"}
```

>>>>

- name: Author 1
  thanks: The authors gratefully acknowledge ...
  affiliation: Department of YYY, University of XXX
- name: Author 2
  affiliation: Department of ZZZ, University of WWW
keywords:
- optimisation
- projection pursuit 
- guided tourr
- visual 
- diagnostics 
- R
abstract: Friedman & Tukey commented on their initial paper on projection pursuit in 1974 that "the  technique  used  for  maximising  the  projection index strongly influences both the statistical and the computational aspects of the procedure." While many projection pursuit indices have been proposed in the literature, few concerns the optimisation procedures. In this paper, we illustrates the diagnostics of optimisers for projection pursuit guided tour via the construction of a data object and producing diagnostic plots. These diagnostics workflow can be applied generally on optimisers to assess its performance. An R package, ferrn, has been created to implement the diagnostics.
bibliography: biblio.bib
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont, bbm, array, booktabs}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 3
editor_options: 
  chunk_output_type: console
---

@@

```{r external, include = FALSE, cache = FALSE}
library(knitr)
library(tidyverse)
library(ferrn)
read_chunk(here::here('scripts/vis-diag.R'))
read_chunk(here::here('scripts/examples.R'))
read_chunk(here::here('scripts/kol_result.R'))
read_chunk(here::here('scripts/implementation.R'))
```

@@

Visualisation has been widely used in exploratory data analysis. Presenting information in a graphical format often allows people to see information they would otherwise not see. This motivates our work of creating plots to diagnose optimisation algorithms with the aim to understand and compare features of different algorithms.

In an optimization problem the goal is to find the best solution within the space of all feasible solutions, which typically is represented by a set of constraints. The problem consists in optimizing an objective function $f: S \rightarrow \mathbb{R}$ with $S \in \mathbb{R}^n$ in a reduced space given by the problem constraints to either minimize or maximize a function.

The problem context of the optimisation that has been worked in this paper is projection pursuit guided tour, which are exploratory data analysis tools that detect interesting structure of high dimensional data through projection on low dimensional space. Optimisation is applied here to search for the low dimensional space that finds the most interesting projection.

The remainder of the paper is organised as follows. 
Section \ref{optim} provides a literature review of optimisation methods, specifically the line search methods. 
Section \ref{tour} reviews projection pursuit guided tour, forms the optimisation problem and its features, and introduces three main existing algorithms.
Section \ref{vis-diag} presents new visual diagnostics designs. Data object is first defined to pinned down the data structure, several diagnostic plots are then defined with sample examples given.
Section \ref{application} shows the application of how the diagnostic plots designed previously can be used to understand and compare different algorithms and how they contribute to modifications that improve the algorithms.
Finally, Section \ref{implementation} describes the R package: ferrn, that implements all the visual diagnostics above.

@@

Given an optimisation problem, two broad approaches find the optimum based on different thinking and hence have its own strength and deficits.
An analytical approach aims to find the optimal solution in a finite number of steps, but the closed-form solution may not be available when the problem starts to become complex. 
An iterative approach, on the other hand, finds the optimum based on the idea of making progressive improvement to the current solution and researchers can decide when to stop if a desirable accuracy has been achieved. However, a potential issue is that it may end up finding a local optimum. In this paper, the main focus is on a particular iterative method:  *line search method* [@fletcher2013practical]. In a simple one-dimensional problem of finding the $x$ that minimises $f(x)$, line search conducts its search via an iterative up as in Equation \ref{eq:line-search}.

@@

where $d^{(j)}$ is the searching direction in iteration $j$, and $\alpha_j$ is the step-size. Theoretical results have demonstrated the global convergence of the algorithm when $\alpha_k$ is chosen by minimising  $f(x^{(j)} + \alpha* d^{(j)})$ [@curry1944method]. However, in practice, this is rarely implemented due to its computational demanding or sometimes the existence of such a minimisation[@fletcher2013practical]. A modified approach is to impose a mandatory decrease in the objective function for each iteration: $f^{(j+1)}> f^{(j)}$ and this approach turns out to be efficient in practical problems despite the lost of guarantee on global convergence.

@@

Modern development of the line search methods focuses on proposing different computation on the searching direction: $d^{(j)}$ and various approximations on the step size: $\alpha_j$ catered for practical optimisation problems. The specific problem context discussed in this paper is projection pursuit guided tour. 
Projection pursuit and guided tour are two separate methods in exploratory data analysis focusing on different aspects: Projection pursuit, coined by @friedman1974projection, detects interesting structures (i.e. clustering, outliers and skewness) in multivariate data via low dimensions projection whilst guided tour is one particular variation in a broader class of data visualisation method, tour.

Let $\mathbf{X}_{n \times p}$ be the data matrix, an n-d projection can be seen as a linear transformation $T: \mathbb{R}^p \mapsto \mathbb{R}^d$ defined by $\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{Y}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis. Define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projection basis $\mathbf{A}$ onto an index value $I$, this function is commonly known as the projection pursuit index (PPI) function, or the index function and is used to measure the "interestingness" of a projection. A number of index functions have been proposed in the literature to detect different data structures, including Legendre index [@friedman1974projection], Hermite index [@hall1989polynomial], natural Hermite index [@cook1993projection], chi-square index [@posse1995projection], LDA index [@lee2005projection] and PDA index [@lee2010projection].

As a general visualisation method, A tour produces animated visualisation of the high dimensional data via rotating low dimension planes. Different tours choose the low dimensional planes differently, for example, a grand tour[@cook2008grand] selects the planes randomly in the high dimensional space and manual tour[@cook1997manual] chooses a sequence of bases that gradually phase in and out one variable to view the contribution of that variable in the visualisation. Guided tour, the main interest of this paper, chooses its projection with the aid of projection pursuit. Given a random start, projection pursuit iteratively finds bases with higher index value and guided tour constructs geodesic interpolation between two consecutive to form a tour path. Mathematical details of the geodesic interpolation can be found in @buja2005computational. Figure \@ref(fig:tour-path) shows a sketch of the tour path consisting of the blue frames produced by the projection pursuit optimisation algorithm iteratively and the white frames, the interpolations between two blue frames. The tour method has been implemented in the `tourr` package in R, available on the Comprehensive R Archive Network at [https://cran.r-project.org/web/packages/tourr/](https://cran.r-project.org/web/packages/tourr/) [@wickham2011tourrpackage].

@@

(ref:tour-path) Each square (frame) represents the projected data with a corresponding basis. Blue frames are found by an optimisation algorithm iteratively whilst the white frames are constructed between two blue frames by geodesic interpolation.

@@

The optimisation problem is stated as follows. Given a randomly generated starting basis $\mathbf{A}_1$, projection pursuit finds the final projection basis $\mathbf{A}_T$ that satisfies the following optimisation problem:

@@

where $I_d$ is the d-dimensional identity matrix and the constraint requires the projection bases $\mathbf{A}$ to be an orthogonal matrices.

Several features of this optimisation are worth noticing. First of all, this is a constraint optimisation problem as the decision variables form the entries of a projection basis, which is required to be orthonormal. It is also likely that the objective function may not be differentiable for a constructed index function and in these cases, gradient-based methods may not work well. Although finding the global maximum is the goal of an optimisation problem, it is also interesting to inspect local maximum in projection pursuit since it could present unexpected interesting projections. Lastly, there is also one computational consideration: the optimisation procedure needs to be fast to compute since the tour animation is played in real-time.

@@

Below we introduce three possible algorithms: `search_better`, `search_better_random`, and `search_geodesic`. The first two are derivative free methods that sample candidate bases in the neighbourhood whilst `search_geodesic` is an analogue of gradient ascent on the projection basis space.

\begin{algorithm}
\SetAlgoLined
  \SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{$\mathbf{A}_{\text{cur}}$, $f$, $\alpha$, $l_{\max}$} 
    \Output{$\mathbf{A}_{l}$}
  initialisation\;
  Set $l = 1$\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ and orthogonalise $\mathbf{A}_{l}$\;
    Compute $I_{l}  = f(\mathbf{A}_{l})$\;
    \If{$I_{l} > I_{\text{cur}}$}{
      \KwRet{$\mathbf{A}_{l}$} \;
      }
    $l = l + 1$\;
  }
  \caption{random search}
  \label{random-search}
\end{algorithm}

`search_better` is a random search device that samples a candidate basis $\mathbf{A}_{l}$ in the neighbourhood of the current basis $\mathbf{A}_{\text{cur}}$ by $\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ where $\alpha$ controls the radius of the sampling neighbourhood and $\mathbf{A}_{\text{rand}}$ is a randomly generated matrix with the same dimension as $\mathbf{A}_{\text{cur}}$. $\mathbf{A}_{l}$ is then orthogonalised to ensure the orthonormal constraint is fulfilled. When a basis is found with index value higher than the current basis $\mathbf{A}_{\text{cur}}$, the search terminates and outputs the basis for guided tour to construct an interpolation path. The next iteration of search begins after adjusting $\alpha$ by a cooling parameter: $\alpha_{j+1} = \alpha_j * \text{cooling}$. The termination condition is when the maximum number of iteration $l_{\max}$ is reached. The algorithm of `search_better` is summarised in Algorithm \ref{random-search}.  A slightly different cooling scheme has been proposed by @posse1995projection to include a halving parameter $c$. Rather than reducing the radius of the searching neighbourhood, $\alpha$, at each iteration, Posse's design only adjust $\alpha$ if the last search takes more than $c$ times to find an accepted basis to avoid the searching space being reduced too fast.

@@

Simulated annealing (`search_better_random`) [@kirkpatrick1983optimization, @bertsimas1993simulated] uses the same sampling process as `search_better` but allow a probabilistic acceptance of a basis with lower index value based on the annealing $T(l)$. Given an initial $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{\log(l + 1)}$. When a candidate basis fails to have an index value larger than the current basis, simulated annealing gives it a second chance to be accepted with probability $$P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}$$ where $I_{(\cdot)}$ denotes the index value of a given basis. This implementation allows the algorithm to jump out of a local maximum and enables a more holistic search of the whole parameter space. This feature is particularly useful when local maximum is presented. The algorithm can be written as replacing line 5-8 of Algorithm \ref{random-search} with Algorithm \ref{simulated_annealing}.

@@

@cook1995grand used a gradient ascent algorithm on the space of the projection bases. In gradient ascent, one first find the direction for improvement via computing the gradient information. In `search_geodesic`, $2n$ bases are first generated in a tiny neighbourhood of the current basis, controlled by the neighbourhood parameter $\delta$. A geodesic is then constructed using the current basis and the one in $2n$ bases with the highest index value. If the neighbourhood parameter $\delta$ is tiny, the geodesic constructed is an analogue of the gradient information in the curved space and works as the searching direction. The next step in gradient ascent is to conduct a line search to find the best improvement along the gradient direction and in `search_geodesic`, this is replaced by optimising the index value along the geodesic direction over an 90 degree angle from $-\pi/4$ to $\pi/4$. The optima $\mathbf{A}_{**}$ is returned for the current iteration if it meets the termination condition on percentage improvement.  The procedure will also terminate if $l_{\max}$ is reached.  Algorithm \ref{search-geodesic} summarises the steps in geodesic search.

@@

To be able to make diagnostics on the optimisers, the algorithms need to populate a data structure with key elements of the algorithm. When the algorithms run, key information regarding the decision variable, objective function and hyper-parameters, needs to be recorded and stored as a data object so that it is ready to be supplied to the plotting functions for diagnostics.

@@

In the optimisation algorithms for projection pursuit, the three main elements to record are 1) projection bases: $\mathbf{A}$, 2) index values: $I$, and 3) State: $S$, which labels the observation with detailed stage in the optimisation. Possible values for `search_better` and `search_better_random` includes `random_search`,  `new_basis`, and `interpolation`. `search_geodesic` has a wider variety that includes `new_basis`, `direction_search`, `best_direction_search`, `best_line_search`, and `interpolation`.

Multiple iterators are also  needed to index the data collected at different levels. $t$ is a unique identifier that prescribes the natural ordering of each observation; $j$ is the counter for each search-and-interpolate iteration, which remains the same within one round and has an increment of one once a new round starts. $l$ is the counter for each search/interpolation, which provides the information of how many basis the algorithm has searched before finding one to return. There are other parameters of interest, depends on the particular problem content and they are denoted as *$V_{p}$*. Two most common examples include $V_1 = \text{method}$, which tags the name of the algorithm used, and $V_2 = \text{alpha}$, the neighbourhood parameter that controls the size in sampling candidate bases.  A matrix notation of the data structure is presented in Equation \ref{eq:data-structure}.

@@

where $T^{\prime} = T + k_{J}+ l_{J+1}$. Note that there is no output in iteration $J + 1$ since the optimiser can't find a better basis and the algorithm terminates. In this notation, final basis found is $A_T$ with the highest index value $I_T$.

@@

The data structure constructed above meets the tidy data principle [@wickham2014tidy] that requires each observation forms a row and each variable forms a column. With tidy data structure, data wrangling and visualisation have been significantly simplified by the well-developed packages like dplyr[@dplyr] and ggplot2[@ggplot2].

The construction of diagnostic plots uses the concept of grammar of graphic [@wickham2010layered] in ggplot2. In grammar of graphic, plots are not defined by its appearance (i.e. boxplot, histogram, scatter plot etc) but by stacked layers. In the construction of diagnostic plots, there are multiple elements one may wants to emphasize and there's no single plot name that would meet the need, the stacked layers concept, on the other hand, allow information to be overlaid on the each other and one can build the plot from scratch as long as the variables have been stored in a dataset.

@@

```{r toy-pca-animated, fig.cap = "(ref:toy-pca-animated)"}
```

(ref:toy-pca-animated) A selected number of frames from the animated PCA plot. With animation, it is easier to track the progression from the start to finish in each algorithm.
